%!TEX root = ../main.tex

\chapter*{Preface}
The 2013 European Conference on Modelling Foundations and Applications (ECMFA 2013) was organised by the Laboratoire d'Informatique de Robotique et de Micro\'{e}lectronique de Montpellier (LIRMM), CNRS, University Montpellier 2, France during July 1-5, 2013.

ECMFA is the key European conference aiming at advancing the techniques and furthering the underlying knowledge related to model-driven engineering. Model-driven engineering (MDE) is a software development approach based on the use of models for the specification, design, analysis, synthesis, deployment, testing and maintenance of complex software systems, aiming to produce high-quality systems at lower costs. In the past eight years, ECMFA has provided an ideal venue for interaction among researchers interested in MDE both from academia and industry. The ninth edition of the conference covered major advances in foundational research and industrial applications of MDE.

In 2013, the Program Committee received 76 abstracts and 51 full paper submissions. From these, 9 Foundations track papers and 6 Applications track papers were accepted for presentation at the conference and publication in these proceedings, resulting in an acceptance rate of 23\% for the foundations track, and an overall acceptance rate of 29.4\%. Papers on all aspects of MDE were received, including topics such as model querying, consistency checking, model transformation, model-based systems engineering and domain-specific modeling. The breadth of topics and the high quality of the results presented in these accepted papers demonstrate the maturity and vibrancy of the field.

The ECMFA 2013 keynote speakers were Martin Gogolla (University of Bremen) and Dierk Steinbach (Cassidian, EADS Deutschland GmbH). We thank them very much for accepting our invitation and for their enlightening talks. We are grateful to our Program Committee members for providing their expertise and quality and timely reviews. Their helpful and constructive feedback to all authors is most appreciated. We thank the ECMFA Conference Steering Committee for their advice and help. We also thank our sponsors and all authors who submitted papers to ECMFA 2013.

\vspace{1cm}
\begin{flushright}\noindent
Pieter Van Gorp \\
Tom Ritter \\
Louis M. Rose \\
$ $ \\
July 2013
\end{flushright}